\name{}: Combining Bottom-up and top-down approaches through graph learning over interaction networks for drug-target-interaction prediction

### 2021-04-09 paper/main.tex
.

\begin{table*}

	\begin{tabular}{|l|p{1cm}|p{1cm}|p{1cm}|p{1cm}|p{1cm}|p{1cm}|p{1cm}|}

		\hline
		&&\multicolumn{3}{c|}{Yamanishi results}&\multicolumn{3}{c|}{BioSnap results}\\
		Approach&Original&Original scheme&\multicolumn{2}{c|}{Protein split}&Original scheme&\multicolumn{2}{c|}{Protein split}\\

		&Splitting scheme&Macro AUC&Macro AUC&Micro $AUC_p$&Macro AUC&Macro AUC&Micro $AUC_p$\\
		\hline
		Naive predictor&DP pairs&$0.85$& --&--&$0.79$&-- &--\\
		DTINet&DP pairs&$0.91$&$0.74$&$0.67$ & --&--&--\\
		DTIGEMS+&DP pairs&$\mathbf{0.93}$& $0.72$& $0.68$ & --&--&--\\
		DTI-CDF&Proteins&$0.85$&$\mathbf{0.85}$&$0.79$& --&--&--\\
		\hline
		DeepDTI&Drugs& --&--&--&$0.88$&$0.76$&$0.70$\\
		DeepDTA&DP pairs& --&--&--&$0.88$&$0.77$&$0.69$\\
		DeepConv-DTI&DP pairs& --&--&--&$0.88$&$0.76$&$0.73$\\
		MolTrans&DP pairs& --&--&--&$0.90$&$0.77$&$0.74$\\
		\name&Proteins&$0.84$&$0.84$&$\mathbf{0.84}$&0.85 & $\mathbf{0.85}$& $\mathbf{0.82}$\\
		\hline
	\end{tabular}
        \caption{\label{tab:comparison} Comparison of \name{} with
          state of the art drug--target interaction prediction methods
          (Yamanishi dataset) and drug--target affinity prediction
          methods (BioSnap dataset); we evaluated the original and the
          protein-based split in a cross-validation}.
\end{table*}

@@

\begin{table*}[!tpb]
  \centering
  \begin{tabular}{|p{2.2cm}|p{0.9cm}|p{0.75cm}|p{0.9cm}|p{0.75cm}|p{0.9cm}|p{0.75cm}|p{0.9cm}|p{0.75cm}|}
    \hline
    &\multicolumn{4}{c|}{(a) STITCH results}& \multicolumn{4}{c|}{(b) Yamanishi results}\\
    \hline
    \name{} results&\multicolumn{4}{c|}{PPI graph}&\multicolumn{4}{c|}{PPI graph}\\
    &\multicolumn{2}{c|}{without}&\multicolumn{2}{c|}{with}&\multicolumn{2}{c|}{without}&\multicolumn{2}{c|}{with}\\
    &Macro AUC&Micro $AUC_p$&Macro AUC&Micro $AUC_p$&Macro AUC&Micro AUC&Macro AUC&Micro $AUC_p$\\
    \hline
    MolPred&$0.69$&$0.65$&$0.69$&$0.67$&$0.66$&$0.67$&$0.66$&$0.64$\\
    \hline
    OntoPred&$0.88$&$0.87$&$0.92$&$0.93$&$0.80$&$0.79$&$0.83$&$0.82$\\
    \hline
    \name{} (MolPred + OntoPred) & $0.89$ & $0.90$&$\mathbf{0.93}$&$\mathbf{0.94}$& $0.83$ & $0.82$&$\mathbf{0.84}$&$\mathbf{0.84}$\\
    \hline
  \end{tabular}
  \caption{\label{tab:Results}Results for \name{} on the STITCH
    and Yamanishi datasets evaluated with 5-fold cross-validation. We 
    call the model using only molecular features
    \textit{MolPred} and the model using only ontology-based
    features \textit{OntoPred}. \name{} combines both types of
    features.}
\end{table*}

@@

We find that the model using ontology-based features
(\textit{OntoPred}) is showing better performance on STITCH compared
to using only molecular features. We also observe that only the model
using ontology-based features results in increased performance when
incorporating the PPI graph. This increase can be observed with
different graph neural network architectures and configurations. While
the {\em GCNConv} and {\em GENConv} architecture already shows some
minor improvement, the use of \textit{ResGraphConv} results in larger
performance improvements.  {\em ResGraphConv} blocks add a large
amount of additional learnable parameters to the network, leading to
more expressive power. To test whether the observed improvement is due
to the number of learnable parameters added or the result of better
exploiting the information about PPIs, we experiment with a graph
model in which all graph convolutional neural layers in the residual
blocks are removed, resulting in a model with similar parameters but
without the ability to use graph-based information. This pruned
network, with no information on the protein--protein interactions,
reached very similar results to the original \textit{OntoPred} model
and showed no improvement.

The improvements when including the graph are only provided by the
\textit{GENConv} graph convolution scheme which includes the
{\em ResGraphConv} blocks; \textit{GCNConv} and other graph
convolutional methods fail to achieve any gain in comparison to the
plain \textit{OntoPred} performance even when combined with the
residual blocks. The discrepancy between {\em GENConv} and other graph
convolutional methods may be the result of numerical instability and
fading signals \cite{}.

@@

% \subsection{Baseline model}
Furthermore, we hypothesize that it may be possible to exploit biases
in drug--target interactions to achieve relatively high prediction
performance without obtaining a biologically meaningful signal.
For example, hub proteins may have a large number of interactions,
or certain drugs interact with many proteins, and preferentially
predicting these interactions may increase predictive performance even
in the absence of any biological features.
To test this hypothesis, we design a ``na\"ive'' baseline model that
predicts the same list of proteins for each drug based only on the
number of known drug--target interactions for a protein. Formally,
given lists $D$ and $P$ of drugs and proteins and a set of known
interactions $\mathcal{I} := \{(d_i, p_i) \}$, we construct an
interaction matrix $M_{int}\in\{0,1\}^{|D|\times|P|}$ with
\begin{equation*}
	M_{ij} = \begin{cases}
		1 & \text{if } (d_i, p_j)\in \mathcal{I}\\
		0 & otherwise
	\end{cases}
\end{equation*}
describing for all drug--protein pairs whether there is a known
interaction or not. We now rank all proteins $p_j\in P$ descending by
their number of drug interactors by summing over the columns of
$M_{ij}$ and ranking these sums:
\begin{equation*}
	f: P \rightarrow \mathbb{N} \text{ with } f:p_j \mapsto \sum_{i=1}^{|D|}M_{ij}
\end{equation*}
Our ``na\"ive'' predictor $P_k$ predicts all drugs to interact with
the top $k$ targets with respect to the introduced ranking:
\begin{equation*}
	P_k: D\times P \rightarrow \{0,1\} \text{ with } P_k: d_i, p_j \mapsto \begin{cases}
		1 & \text{ if }p_j \in Top_{k}(P)\\
		0 & otherwise
	\end{cases}
\end{equation*}
with the only hyperparameter $k$.

The prediction $P_k$ is not dependent on the drug $d_i$ and will
predict the same ranked list of drugs for all proteins; consequently,
this na\"ive predictor does not rely on any biological features and
will not predict any novel information about interactions between
drugs and proteins; the na\"ive predictor only exploits imbalances in
the evaluation set to make predictions that may perform
well. % but allows us to estimate the effect of potential
% biases in our evaluation dataset.
% Also note the possibility for a similar predictor by calculating the top $k$ interacting drugs, respectively.

The way in which we formulated the na\"ive predictor, it is not
applicable for a protein split cross-validation as the number of
interactions for each protein in the validation set is unknown.  We
apply this naive predictor on both the STITCH and Yamanishi datasets,
using the full datasets as well as a 5-fold cross-validation over
drugs and over drug--protein pairs to compare the prediction results
directly to \name{}.  For each fold, we gradually increase $k$ to
determine the best performance for each fold. Using the full dataset,
drug--target split, and drug split, we obtain the following MacroAUC
results\todo[inline]{Put in Results table, then remove from here}: for
the STITCH database, we obtain a performance of $0.76$ on the whole
dataset, $0.70$ for the drug--target pairs and $0.73$ in case of the
drug splitting scheme; on the Yamanishi dataset, we obtain MacroAUC
scores of $0.88$, $0.84$ and $0.85$, for the total dataset,
drug--target pair and drug split, respectively.  The na\"ive predictor
shows higher performance on the Yamanishi dataset than on STITCH, and
a substantial gain in comparison to an expected random predictor on
both datasets. In the following, we utilize this nai\"ve predictor as
baseline to compare its performance to state of the art models and
\name{}.

@@

The results of our experiments are summarized in part (c) of Table
\ref{tab:Results}\todo{Check after restructure}; we calculated the
performance of all compared methods over their original splitting
scheme and over a protein split. We find that there is a large
difference in performance when evaluating over a drug--target pair
split compared to a protein split, with generally higher performance
achieved when using the drug--target pair split.  Second, when
evaluating the same methods over a protein split, we find a
substantial performance difference in comparison to the splitting
scheme used in the original evaluation of each method. DTI-CDF was
originally evaluated on all three splitting schemes underlining this
point \cite{}. While \name{} provides comparable performance to the
na\"ive predictor and DTI-CDF in terms of MacroAUC, it yields
considerably better results with respect to $\textrm{MicroAUC}_p$. We
also find that methods that are trained using a protein split
generally result in higher $\textrm{MicroAUC}_p$ than methods trained
using a drug--target pair split.

As the difference in performance with different splitting schemes is
quite large, we further evaluated additional drug--target interaction
and drug--target affinity prediction methods that were trained and
evaluation on other datasets.  Following the results of MolTrans
\cite{MolTrans2020}, we reevaluated DeepDTI \cite{DeepDTI2017}, DeepDTA
\cite{DeepDTA2018}, DeepConv-DTI \cite{DeepConvDTI2019} and MolTrans itself on BioSnap\todo{What
  is that? Is it described in methods?}  \cite{} over a protein split,
as shown in part (d) of Table \ref{tab:Results}\todo{check table}. The
authors of MolTrans evaluated all considered methods over the
drug--target pair and the protein split; we were able to reproduce the
results (Table \ref{tab:Results} (d))\todo{check table}, showing a
substantial difference based on the splitting scheme. We additionally
computed the $MicroAUC_p$ score for all considered methods, leading to
evaluation results that are comparable\todo{are the datasets
  comparable?} to the ones obtained on the Yaminishi dataset.

\todo[inline]{ add \name{} model to table to allow
  DIRECT comparison:}
The results of the analysis are summarized in the upper part of Figure
\ref{tab:Results}.

We find that...\todo[inline]{summarize key findings from Table.}

\todo[inline]{Replace/merge table:}

%As the baseline predictor is based on the second dataset bias i.e. the existence of hub proteins with significantly more interactions within the dataset, this may hint on a more severe skew within the Yamanishi benchmark.

>>>>

@@

\begin{table}[ht]
  \centering
  \begin{tabular}{|p{1.1cm}|p{0.6cm}|p{0.7cm}|p{0.6cm}|p{0.7cm}|p{0.6cm}|p{0.7cm}|p{0.6cm}|p{0.7cm}|}
    \hline
    &\multicolumn{4}{c|}{(a) STITCH results}& \multicolumn{4}{c|}{(b) Yamanishi results}\\
    \hline
    \name{} results&\multicolumn{4}{c|}{PPI graph}&\multicolumn{4}{c|}{PPI graph}\\
    &\multicolumn{2}{c|}{without}&\multicolumn{2}{c|}{with}&\multicolumn{2}{c|}{without}&\multicolumn{2}{c|}{with}\\
    &Macro AUC&Micro $AUC_p$&Macro AUC&Micro $AUC_p$&Macro AUC&Micro AUC&Macro AUC&Micro $AUC_p$\\
    \hline
    MolPred&$0.69$&$0.65$&$0.69$&$0.67$&$0.66$&$0.67$&$0.66$&$0.64$\\
    \hline
    OntoPred&$0.88$&$0.87$&$0.92$&$0.93$&$0.80$&$0.79$&$0.83$&$0.82$\\
    \hline
    \name{} & $0.89$ & $0.90$&$\mathbf{0.93}$&$\mathbf{0.94}$& $0.83$ & $0.82$&$\mathbf{0.84}$&$\mathbf{0.84}$\\
    \hline
  \end{tabular}
  \caption{\label{tab:Results}Results for \name{} on the STITCH
    and Yamanishi datasets evaluated with 5-fold cross-validation. We 
    call the model using only molecular features
    \textit{MolPred} and the model using only ontology-based
    features \textit{OntoPred}. \name{} combines both types of
    features.}
\end{table}

@@

We find that the model using ontology-based features
(\textit{OntoPred}) is showing better performance on STITCH compared
to using only molecular features. We also observe that only the model
using ontology-based features results in increased performance when
incorporating the PPI graph. This increase can be observed with
different graph neural network architectures and configurations
(Supplementary Table X\todo{Add to Supplement!}). While the {\em
  GCNConv} and {\em GENConv} architecture already shows some minor
improvement, the use of \textit{ResGraphConv} results in larger
performance improvements.  {\em ResGraphConv} blocks add a large
amount of additional learnable parameters to the network, leading to
more expressive power. To test whether the observed improvement is due
to the number of learnable parameters added or the result of better
exploiting the information about PPIs, we experiment with a graph
model in which all graph convolutional neural layers in the residual
blocks are removed, resulting in a model with similar parameters but
without the ability to use graph-based information. This pruned
network, with no information on the protein--protein interactions,
reached very similar results to the original \textit{OntoPred} model
and showed no improvement (Supplementary Table X\todo{Add!}).

The improvements when including the graph are only provided by the
\textit{GENConv} graph convolution scheme which includes the
{\em ResGraphConv} blocks; \textit{GCNConv} and other graph
convolutional methods fail to achieve any gain in comparison to the
plain \textit{OntoPred} performance even when combined with the
residual blocks. The discrepancy between {\em GENConv} and other graph
convolutional methods may be the result of numerical instability and
fading signals \cite{}\todo{Add citation to original paper}.

@@

% \subsection{Baseline model}
Furthermore, we hypothesize that it may be possible for a machine
learning model to exploit biases in drug--target interaction data to
achieve relatively high prediction performance without obtaining a
biologically meaningful signal.  For example, hub proteins may have a
large number of interactions, or certain drugs interact with many
proteins, and preferentially predicting these interactions may
increase predictive performance even in the absence of any biological
features.  To test this hypothesis, we design a ``na\"ive'' baseline
model that predicts the same list of proteins for each drug based only
on the number of known drug--target interactions for a
protein. Specifically, given lists $D$ and $P$ of drugs and proteins
and a set of known interactions $\mathcal{I} := \{(d_i, p_i) \}$, we
construct an interaction matrix $M_{int}\in\{0,1\}^{|D|\times|P|}$
with
\begin{equation*}
	M_{ij} = \begin{cases}
		1 & \text{if } (d_i, p_j)\in \mathcal{I}\\
		0 & otherwise
	\end{cases}
\end{equation*}
describing for all drug--protein pairs whether there is a known
interaction or not. We now rank all proteins $p_j\in P$ descending by
their number of drug interactors by summing over the columns of
$M_{ij}$ and ranking these sums:
\begin{equation*}
	f: P \rightarrow \mathbb{N} \text{ with } f:p_j \mapsto \sum_{i=1}^{|D|}M_{ij}
\end{equation*}
Our ``na\"ive'' predictor $P_k$ predicts all drugs to interact with
the top $k$ targets with respect to the introduced ranking:
\begin{equation*}
	P_k: D\times P \rightarrow \{0,1\} \text{ with } P_k: d_i, p_j \mapsto \begin{cases}
		1 & \text{ if }p_j \in Top_{k}(P)\\
		0 & otherwise
	\end{cases}
\end{equation*}
with the only hyperparameter $k$.

The prediction $P_k$ is not dependent on the drug $d_i$ and will
predict the same ranked list of drugs for all proteins; consequently,
this na\"ive predictor does not rely on any biological features and
will not predict any novel information about interactions between
drugs and proteins; the na\"ive predictor only exploits imbalances in
the evaluation set to make predictions that may perform well.  The way
in which we formulated the na\"ive predictor, it is not applicable for
a protein split cross-validation as the number of interactions for
each protein in the validation set is unknown.

We apply this naive predictor on both the STITCH and Yamanishi
datasets, using the full datasets as well as a 5-fold cross-validation
over drugs and over drug--protein pairs to compare the prediction
results directly to \name{}.  For each fold, we gradually increase $k$
to determine the best performance for each fold. Using the full
dataset, drug--target split, and drug split, we obtain the following
MacroAUC results: for the STITCH database, we obtain a performance of
$0.76$ on the whole dataset, $0.70$ for the drug--target pairs and
$0.73$ in case of the drug splitting scheme; on the Yamanishi dataset,
we obtain MacroAUC scores of $0.88$, $0.84$ and $0.85$, for the total
dataset, drug--target pair and drug split, respectively\todo{In the
  Table is says 0.85 on Yaminishi for DP pair split, here it says 0.84
  -- which is correct? Need fix}.  The na\"ive
predictor shows higher performance on the Yamanishi dataset than on
STITCH, and a substantial gain in comparison to an expected random
predictor on both datasets. In the following, we utilize this nai\"ve
predictor as baseline to compare its performance to state of the art
models and \name{}.

@@

\begin{table}[ht]
  \centering
  \begin{tabular}{|l|p{1cm}|p{1cm}|p{1cm}|p{1cm}|}
    \hline
    Approach&Original&Original scheme&\multicolumn{2}{c|}{Protein split}\\

    &Splitting scheme&Macro AUC&Macro AUC&Micro $AUC_p$\\
    \hline
    Naive predictor&DP pairs&$0.85$& --&--\\
    DTINet&DP pairs&$0.91$&$0.74$&$0.67$ \\
    DTIGEMS+&DP pairs&$\mathbf{0.93}$& $0.72$& $0.68$ \\
    DTI-CDF&Proteins&$0.85$&$\mathbf{0.85}$&$0.79$\\
    \name&Proteins&$0.84$&$0.84$&$\mathbf{0.84}$\\
    \hline
  \end{tabular}
  \caption{\label{tab:comparison1} Comparison of \name{} with
    state of the art drug--target interaction prediction methods on
    the Yamanishi dataset; we evaluated the original and the
    protein-based split in a cross-validation}.
\end{table}

The results of our experiments are summarized in Table
\ref{tab:comparison1}; we calculated the performance of all compared
methods over their original splitting scheme and over a protein
split. We find that there is a large difference in performance when
evaluating over a drug--target pair split compared to a protein split,
with generally higher performance achieved when using the drug--target
pair split.  Second, when evaluating the same methods over a protein
split, we find a substantial performance difference in comparison to
the splitting scheme used in the original evaluation of each
method. DTI-CDF was originally evaluated on all three splitting
schemes underlining this point \cite{}. While \name{} provides
comparable performance to the na\"ive predictor and DTI-CDF in terms
of MacroAUC, it yields considerably better results with respect to
$\textrm{MicroAUC}_p$. We also find that methods that are trained
using a protein split generally result in higher $\textrm{MicroAUC}_p$
than methods trained using a drug--target pair split, indicating that
they may generalize better to unseen protein targets whereas methods
trained on a drug--target split potentially exploit hidden biases and
therefore generalize less well.

\begin{table}[ht]
  \centering
  \begin{tabular}{|l|p{1cm}|p{1cm}|p{1cm}|p{1cm}|}
    \hline
    Approach&Original&Original scheme&\multicolumn{2}{c|}{Protein split}\\

    &Splitting scheme&Macro AUC&Macro AUC&Micro $AUC_p$\\
    \hline
    Naive predictor&DP pairs&$0.79$&-- &--\\
    DeepDTI&Drugs&$0.88$&$0.76$&$0.70$\\
    DeepDTA&DP pairs&$0.88$&$0.77$&$0.69$\\
    DeepConv-DTI&DP pairs&$0.88$&$0.76$&$0.73$\\
    MolTrans&DP pairs&$0.90$&$0.77$&$0.74$\\
    \name&Proteins&0.85 & $\mathbf{0.85}$& $\mathbf{0.82}$\\
    \hline
  \end{tabular}
  \caption{\label{tab:comparison2} Comparison of \name{} with
    state of the art drug--target interaction prediction methods
    on the BioSnap dataset; we evaluated the original and the
    protein-based split in a cross-validation}.
\end{table}

As the difference in performance with different splitting schemes is
quite large, we further evaluated additional drug--target interaction
and drug--target affinity prediction methods that were trained and
evaluation on other datasets.  Following the results of MolTrans
\cite{MolTrans2020}, we reevaluated DeepDTI \cite{DeepDTI2017},
DeepDTA \cite{DeepDTA2018}, DeepConv-DTI \cite{DeepConvDTI2019}, and
MolTrans itself on the BioSnap dataset \cite{} and compared it to our
``na\"ive'' predictor as well as \name{} (see Table
\ref{tab:comparison2}).  MolTrans was evaluated over the drug--target
pair and the protein split; we were able to reproduce the MolTrans
results (Table \ref{tab:comparison2}), showing a substantial
difference based on the splitting scheme. We additionally computed the
$MicroAUC_p$ score for all considered methods, leading to similar 
results as observed on the Yaminishi dataset.

In all our experiments, \name{} has the highest Micro $AUC_p$,
demonstrating that \name{} can improve over other methods in the task
of identifying drugs that target a specific protein. \name{}

(i.e., higher MacroAUC
when evaluated on a drug--target split compared to a protein split;
large differences in macro- and micro measures)
